Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2022-01-28 Thesis/02-chap2.Rmd
modify some words

Model diagnostics is the part of data analysis, preceded by the fit of a model, whose primary objectives are to examine the goodness of fit and reveal potential violations of assumptions. 
In these diagnostics, though numeric summaries are available and some are endorsed by finite or asymptotic properties, graphic representation of data is preferred or at least needed, due to its intuitiveness and the possibility to provide unexpected discoveries that may be abstract and could not be quantified algorithmically.

However, unlike confirmatory data analysis which is built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding a separation between gene groups in a two-dimensional projection from a linear discriminant analysis where there is no difference in the expression levels between the gene groups [@roy_chowdhury_using_2015].

@@

The size of the collection $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ depends on the size of the set $I$. If one can define $I$ comprehensively, i.e, pre-specify all the visual discoverable features, the validity issue will be solved. Unfortunately, to our knowledge, there is no such a way to list all visual features. In linear regression diagnostics, possible visual features of a residual plot may be outliers, shapes and clusters. But these are very vague features which does not cover the full list of visual features.

Similarly, @wilkinson_graph-theoretic_2005 proposed the work called graph theoretic scagnostics, which adopted the idea of "scagnostics" - scatter plot diagnostics from . It includes 9 computable scagnostics measures defined on planar proximity graphs: "Outlying", "Convex", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy" and "Striated" which attempts to describe outliers, shape, density, trend and coherence of the data. This approach is inspiring but it does not contain the complete list of visual discoverable features. It is possible that such a list will never be complete as suggested in @buja_statistical_2009.

@@

```{r nonlinearityheter, fig.cap="test"}
library(ggplot2)
set.seed(10086)
cm <- visage::cubic_model(a = 0, b = 100, c = 1)

dat <- cm$gen(1000, fit_model = TRUE)
cm$plot(dat) +
  ggtitle("Residuals against fitted values plot of a classical linear regression model") +
  theme_light()
```

>>>>

Model diagnostics is the part of data analysis, preceded by the fit of a model, whose primary objectives are to examine the goodness of fit and reveal potential violations of model assumptions. 
In these diagnostics, though numeric summaries are mostly available and some are even endorsed by finite or asymptotic properties, graphic representation of data is still preferred, or at least needed, due to its intuitiveness and the possibility to provide unexpected discoveries which may be abstract and
unquantifiable.

However, unlike confirmatory data analysis built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding a separation between gene groups in a two-dimensional projection from a linear discriminant analysis where there is no difference in the expression levels between the gene groups [@roy_chowdhury_using_2015].

@@

The size of the collection $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ depends on the size of the set $I$. Thus, if one can define $I$ comprehensively, i.e, pre-specify all the visual discoverable features, the validity issue will be solved. Unfortunately, to our knowledge, there is no such a way to list all visual features. In linear regression diagnostics, possible visual features of a residual plot may be outliers, shapes and clusters. But this is an incomplete list which does not enumerate all the visual features.

Similarly, @wilkinson_graph-theoretic_2005 proposed the work called graph theoretic scagnostics, which adopted the idea of "scagnostics" - scatter plot diagnostics from (can't find the 1984 citation). It includes 9 computable scagnostics measures defined on planar proximity graphs: "Outlying", "Convex", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy" and "Striated" which attempts to describe outliers, shape, density, trend and coherence of the data. This approach is inspiring but it still does not give the complete list of visual discoverable features. In fact, it is possible that such a list will never be complete as suggested in @buja_statistical_2009.

@@

```{r nonlinearityheter, fig.cap="test"}
library(tidyverse)

set.seed(10086)
cm <- visage::cubic_model(a = 0, b = 100, c = 1)

cm$gen(1000, fit_model = TRUE) %>%
cm$plot() +
  ggtitle("Residuals against fitted values plot of a classical linear regression model") +
  theme_light()
```
